---
title: "Examining Predictors of Car Crashes Caused by Alcohol Using Logistic Regression"
author: "Akira Di Sandro, Sofia Fasullo, Amy Solano"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, eval = T, warning = F, message = F)

# get rid of scientific notation
options(scipen = 999)

# set working directory
# setwd("~/Library/Mobile Documents/com~apple~CloudDocs/Documents/MUSA/Fall24/MUSA5000_Stats/MUSA5000_Stats/hw04")

# load packages
library(tidyverse)
library(sf)
library(spatstat)
library(ggplot2)
library(proxy)
library(rnaturalearth)

```

# Introduction

```{r load_data}
farmers_market <- st_read("HW 4/Philadelphia_Farmers_Markets201302.shp", quiet = T)

philly_zips <- st_read("HW 4/Philadelphia_ZipCodes.shp", quiet = T)

philly <- st_read("HW 4/Philadelphia.shp", quiet = T)

# load hydro data for philly
# source: https://opendataphilly.org/datasets/hydrology/
hydro <- st_read("https://services.arcgis.com/fLeGjb7u4uXqeF9q/arcgis/rest/services/Hydrographic_Features_Poly/FeatureServer/1/query?outFields=*&where=1%3D1&f=geojson", quiet = T) %>% 
    st_transform(crs = st_crs(farmers_market))

```



Are Farmers Markets in Philadelphia clustered?

```{r FMmap, fig.cap="Fig 1. Philadelphia Farmers Markets}
# create map
ggplot() +
  
  # philly data
  geom_sf(data = philly, fill = "transparent", color = "lightgrey", lwd = 2) +
  
  # philly_zips data
  geom_sf(data = philly_zips, fill = "darkgrey", color = "lightgrey", lwd = 0.5) + 
  
  # farmers market data
  geom_sf(data = farmers_market) +
  
  # hydro data
  geom_sf(data = hydro, fill = "#96dbe3", color = "transparent") + 
  
  # map limits
  coord_sf(xlim = c(2660000, 2749276),
           ylim = c(208915, 310000)) +
  
  labs(title = "Philadelphia Farmers Market Locations") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5),
        panel.background = element_rect(fill = "#555555"))
  

```


Goal: determine if farmers markets in Philadelphia are clustered, randomly distributed, or dispersed using nearest neighbors analysis and k-function analysis.

# Methods




# Results

```{r convert2PPP}
# get coords for farmers markets
FM_pts <- data.frame(farmers_market %>% 
  st_coordinates())

# got a warning about duplicated points, and figured out that row 25 and 41 are duplicates
# checked the farmer_market data and they are in fact both at RTM
# removing the second point to have all unique points
FM_pts_nodup <- FM_pts[-41,]

# boundary
philly_bound <- as.owin(philly)

# define ppp object
FM_ppp <- ppp(FM_pts_nodup$X, FM_pts_nodup$Y, window = philly_bound)

```


## Nearest Neighbors Analysis (NNA)

```{r NNA}
# Computes the distance from each point to its nearest neighbor in a point pattern.
nnd <- nndist.ppp(FM_ppp)

# calculate Mean Observed Distance, Mean Expected Distance and SE.
MeanObsDist <- mean(nnd)
MeanExpDist <- 0.5 / sqrt(nrow(farmers_market) / area.owin(philly_bound))
SE <- 0.26136 / sqrt(nrow(farmers_market)^2 / area.owin(philly_bound))

# calculate z-statistic
zscore <- (MeanObsDist - MeanExpDist)/SE                     # z = -3.106
pval <- ifelse(zscore > 0, 1 - pnorm(zscore), pnorm(zscore)) # p = 0.0009

# calculating the NNI
NNI <- MeanObsDist / MeanExpDist    # NNI = 0.794

```

The Nearest Neighbor Analysis results in a nearest neighbor index (NNI) of 0.794 with a z-score fo -3.106 and p < 0.001 (p = 0.0009). Though the NNI is closer to 1 than it is to zero, our z-score is less than -1.96, which means the average observed distance is smaller than the average expected distance, meaning we see significant clustering in the distribution of farmers markets across Philadelphia. We can reject the null hypothesis in favor for the alternative hypothesis stating that the observed point pattern is not random.


```{r NNA_spatialEco, echo=FALSE, results='hide'}
spatialEco::nni(farmers_market[-41,], win = "hull")
```
We also conducted the NNA with the nni() function in the spatialEco package. This method gave us a higher NNI of 1.136 with a z-score of 2.028 and a p-value of 0.043. In this case, NNI > 1 and z-score > 1.96, so we would reject the null hypothesis to favor the alternative hypothesis that we observe statistically significant dispersion. 

We expect that this inconsistency between the two methods comes from the fact that our latter method does not accept a boundary argument and assumes a convex hull for study area. As a result, this method used an expected mean distance of 2,797 instead of 4,001 that we calculated for the former method. Since the question we are trying to answer with this question is whether farmers markets are clustered in Philadelphia, our study area boundary is central to that question. Since the former method takes into consideration study area, we focus on the results of that method.


```{r kerneldensitymap, fig.cap="Fig 2. Kernel Density Plot of Philadelphia Farmers Markets}
# prof did this with the base plot functions
plot(density(FM_ppp),
     main = "Kernel Density Plot of Farmers Markets in Philadelphia")
contour(density(FM_ppp), add = T)
plot(FM_ppp, add = T)

```


## K-Function Analysis

```{r kFunction, echo=FALSE, results='hide'}
# find max distance between points
max.distance <- max(proxy::dist(data.frame(cbind(x=FM_pts_nodup$X, y=FM_pts_nodup$Y)),
                                data.frame(cbind(x=FM_pts_nodup$X, y=FM_pts_nodup$Y)),
                                method="euclidean"))
# max.distance  # 56698.31

# approximate 56,698.31 to be about 57,000 and dived that by 2 to get rmax
rmax <- 57000 / 2 # 28500

# define khat
khat <-Kest(FM_ppp, 
            rmax = rmax, 
            correction = "Ripley")

# calculate Ripley's Simulation Confidence Envelopes
Kenv <- spatstat.explore::envelope(FM_ppp,
                                   fun = "Kest", 
                                   rmax = rmax,
                                   nsim = 99,
                                   nrank = 1) 

```



```{r plotKFunction}
# plot Ripley's K-function with 90% simulation envelopes
plot(Kenv,
     xlab = "r", ylab = "Khat(r)", 
     cex.lab = 1.6, cex.axis = 1.5,
     main = "Ripley's K-Function with Confidence Envelopes",
     cex.main = 1.5, lwd = 2)

```

```{r LFunction, echo=FALSE, results='hide'}
# L Function
lhat <- Lest(FM_ppp,
             rmax = rmax,
             correction = "Ripley")

# L Function with 90% envelopes
Lenv <- spatstat.explore::envelope(FM_ppp,
                                   fun = "Lest",
                                   rmax = rmax,
                                   nsim = 99,
                                   nrank = 1)

```


```{r LFunction_prettierplot}
L2 <- Lenv 
L2$obs <- L2$obs-L2$r
L2$theo <- L2$theo-L2$r
L2$lo <- L2$lo-L2$r
L2$hi <- L2$hi-L2$r

plot(L2,
     xlab = "r", ylab = "Lhat(r)", 
     cex.lab = 1.6, cex.axis = 1.5,
     main = "Ripley's L-function with Confidence Envelopes",
     cex.main = 1.5, lwd = 2)
```




```{r FM_map_income}
# create map
ggplot() +
  
  # philly data
  geom_sf(data = philly, fill = "transparent", color = "lightgrey", lwd = 2) +
  
  # philly_zips data
  geom_sf(data = philly_zips, aes(fill = MedIncome), color = "lightgrey", lwd = 0.5) + 
  
  # farmers market data
  geom_sf(data = farmers_market) +
  
  # hydro data
  geom_sf(data = hydro, fill = "#96dbe3", color = "transparent") + 
  
  # map limits
  coord_sf(xlim = c(2660000, 2749276),
           ylim = c(208915, 310000)) +
  
  # change themes
  scale_fill_gradient(low = "#742C14FF",
                      high = "#F0D77BFF",
                      breaks = c(0, 15000, 30000, 45000, 60000),
                      limits = c(0,60000),
                      labels = c("$0", "$15k", "$30k", "$45k", "$60k")) +
  
  labs(title = "Philadelphia Farmers Market Locations and Median HH Income by ZIP code",
       fill = "Med. HH Inc.") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5),
        panel.background = element_rect(fill = "#555555"),
        legend.position = c(.95, .05),
        legend.justification = c("right", "bottom"),
        legend.margin = margin(3, 3, 3, 3))
  

```


