---
title: "Examining Predictors of Car Crashes Caused by Alcohol Using Logistic Regression"
author: "Akira Di Sandro, Sofia Fasullo, Amy Solano"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, eval = T, warning = F, message = F)

# get rid of scientific notation
options(scipen = 999)

# set working directory
# setwd("~/Library/Mobile Documents/com~apple~CloudDocs/Documents/MUSA/Fall24/MUSA5000_Stats/MUSA5000_Stats/hw04")

# load packages
library(tidyverse)
library(sf)
library(spatstat)
library(ggplot2)
library(proxy)


library(graphics)
library(maptools)
library(sp)
library(fossil)
library(spatial)
library(adehabitatHR)
library(gdata)
library(raster)
library(spatialEco)
library(spatstat.model)
library(spatstat.explore)

```


```{r load_data}
farmers_market <- st_read("HW 4/Philadelphia_Farmers_Markets201302.shp", quiet = T)

philly_zips <- st_read("HW 4/Philadelphia_ZipCodes.shp", quiet = T)

philly <- st_read("HW 4/Philadelphia.shp", quiet = T)

```


I first want to take a look at what the data looks like

```{r quickmaps}
ggplot() +
  
  # philly data
  geom_sf(data = philly) +
  
  # philly_zips data
  geom_sf(data = philly_zips) + 
  
  # farmers market data
  geom_sf(data = farmers_market) +
  
  theme_void()
  
```


Goal: determine if farmers markets in Philadelphia are clustered, randomly distributed, or dispersed using nearest neighbors analysis and k-function analysis.

```{r convert2PPP}
# get coords for farmers markets
FM_pts <- data.frame(farmers_market %>% 
  st_coordinates())

# got a warning about duplicated points, and figured out that row 25 and 41 are duplicates
# checked the farmer_market data and they are in fact both at RTM
# removing the second point to have all unique points
FM_pts_nodup <- FM_pts[-41,]

# boundary
philly_bound <- as.owin(philly)

# define ppp object
FM_ppp <- ppp(FM_pts_nodup$X, FM_pts_nodup$Y, window = philly_bound)

```


# Nearest Neighbors Analysis

```{r NNA}
# Computes the distance from each point to its nearest neighbor in a point pattern.
nnd <- nndist.ppp(FM_ppp)

# calculate Mean Observed Distance, Mean Expected Distance and SE.
MeanObsDist <- mean(nnd)
MeanExpDist <- 0.5 / sqrt(nrow(farmers_market) / area.owin(philly_bound))
SE <- 0.26136 / sqrt(nrow(farmers_market)^2 / area.owin(philly_bound))

# calculate z-statistic
zscore <- (MeanObsDist - MeanExpDist)/SE                     # z = -3.106
pval <- ifelse(zscore > 0, 1 - pnorm(zscore), pnorm(zscore)) # p = 0.0004

# calculating the NNI
NNI <- MeanObsDist / MeanExpDist    # NNI = 0.794

```


```{r NNA_spatialEco}
spatialEco::nni(farmers_market[-41,], win = "hull")
```
The spatialEco method gives us an even higher NNI (more dispersed) with a p-value of 0.04


Possibly add a kernel density plot of farmers markets for a nice visual?
TODO!!
can make this prettier in ggplot

```{r}
# prof did this with the base plot functions
plot(density(FM_ppp))
contour(density(FM_ppp), add = T)
plot(FM_ppp, add = T)

```


# K-function Analysis

```{r kFunction}
# find max distance between points
max.distance <- max(proxy::dist(data.frame(cbind(x=FM_pts_nodup$X, y=FM_pts_nodup$Y)),
                                data.frame(cbind(x=FM_pts_nodup$X, y=FM_pts_nodup$Y)),
                                method="euclidean"))
# max.distance  # 56698.31

# approximate 56,698.31 to be about 57,000 and dived that by 2 to get rmax
rmax <- 57000 / 2 # 28500

# define khat
khat <-Kest(FM_ppp, 
            rmax = rmax, 
            correction = "Ripley")

# calculate Ripley's Simulation Confidence Envelopes
Kenv <- spatstat.explore::envelope(FM_ppp,
                                   fun = "Kest", 
                                   rmax = rmax,
                                   nsim = 99,
                                   nrank = 1) 

```

TODO!!
make this in ggplot to make it prettier

```{r plotKFunction}
# plot Ripley's K-function with Edge correction
plot(khat$r, khat$iso, 
     xlab = "r", ylab = "Ripley's K",
     main = "Ripley's Estimated K-Function",
     cex.lab = 1.6, cex.axis = 1.5, cex.main = 1.5,
     lty = 1, lwd = 2)
lines(khat$r, khat$theo, 
      lty = 8, lwd = 2) 

# plot Ripley's K-function with 90% simulation envelopes
plot(Kenv,
     xlab = "r", ylab = "Khat(r)", 
     cex.lab = 1.6, cex.axis = 1.5,
     main = "Ripley's Khat with Confidence Envelopes",
     cex.main = 1.5, lwd = 2)

```

```{r LFunction}
# L Function
lhat <- Lest(FM_ppp,
             rmax = rmax,
             correction = "Ripley")

# L Function with 90% envelopes
Lenv <- spatstat.explore::envelope(FM_ppp,
                                   fun = "Lest",
                                   rmax = rmax,
                                   nsim = 99,
                                   nrank = 1)

```


```{r plotLFunction}
# plot Ripley's L-function
plot(lhat$r, lhat$iso - lhat$r, 
     xlab = "r", ylab = "Ripley's L",
     cex.lab = 1.6, cex.axis = 1.5, cex.main = 1.5, 
     lty = 1, lwd = 2, 
     main="Ripley's Estimated L-Function") 
lines(lhat$r, lhat$theo - lhat$r, 
      lty = 8, lwd = 2) 

# plot Ripley's L-function with 90% simulation envelopes
plot(Lenv,
     xlab = "r", ylab = "Lhat(r)", 
     cex.lab = 1.6, cex.axis = 1.5,
     main = "Ripley's L-function with Confidence Envelopes",
     cex.main = 1.5, lwd = 2)
```

```{r LFunction_prettierplot}
L2 <- Lenv 
L2$obs <- L2$obs-L2$r
L2$theo <- L2$theo-L2$r
L2$lo <- L2$lo-L2$r
L2$hi <- L2$hi-L2$r

plot(L2,
     xlab = "r", ylab = "Lhat(r)", 
     cex.lab = 1.6, cex.axis = 1.5,
     main = "Ripley's L-function with Confidence Envelopes",
     cex.main = 1.5, lwd = 2, legend = F)
```





TODO!!
present a map of the farmers markets overlaid on a map of median household income at the zip code level