---
title: "Predicting Median House Value in Philadelphia"
subtitle: "Comparing OLS, Spatial Lag, Spatial Error Models"
author: "Akira Di Sandro, Sofia Fasullo, Amy Solano"
date: "2024-10-30"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
  pdf_document:
    toc: true
bibliography: ../hw01/HW1_ref.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, eval = T, warning = F, message = F)

# get rid of scientific notation
options(scipen = 999)

# load appropriate packages
library(tidyverse)
library(sf)
library(FNN)
library(grid)
library(gridExtra)
library(kableExtra)
library(ggcorrplot)
library(rmarkdown)
library(pander)
library(Hmisc)
library(car)
library(DAAG)
library(MASS)
library(spdep)
library(spgwr)
library(tmap)
library(spatialreg)
library(whitestrap)
library(lmtest)
library(tseries)
```

# Introduction

This study looks to understand whether several demographic and housing factors can be used to reasonably predict the median house value of a given census block group area in the city of Philadelphia, Pennsylvania. These variables include the percent of houses that are vacant on the block (`PCTVACANT`), the percent of residents with at least a bachelor’s degree (`PCTBACHMOR`), the number of households living in poverty (`NBELPOV100`), and the percent of homes that are single-family detached units (`PCTSINGLES`). We used 2000 Decennial Census data for this study [@acs2000].

With each of the following variables that we will regress median household value upon, a correlation between them and the median household value can be inferred:

-   The presence of vacant buildings on a block would decrease the value of the houses around it, and the vacant buildings themselves would be a lower value. Therefore, the percentage of vacant houses would be negatively correlated with median house value.

-   Typically, it is assumed that people with bachelor’s degrees are more likely to land higher-paying jobs than people without, and therefore can afford more expensive homes. The percentage of people with bachelor degrees would be positively correlated with median house value.

-   If a household is living below the poverty line, they are unlikely to be able to afford expensive living. Therefore the number of households living below the poverty line would be negatively correlated with median house value.

-   Owning a single-family detached house is culturally considered “the American dream,” because one does not have to share walls or amenities with other people. The percentage of single-family units would be positively correlated with the median house value.

```{r load_data, echo=F, include=F}
regdata <- read.csv("../hw01/RegressionData.csv")
regshp <- st_read("data/HW02_RegressionData.shp")

```

```{r queen_neighbors}
queen <- poly2nb(regshp, row.names = regshp$POLY_ID)

# summary(queen)
```

```{r globalMoransI}
queenlist <- nb2listw(queen, style = 'W')

moran(regshp$LNMEDHHINC, queenlist, n=length(queenlist$neighbours), S0=Szero(queenlist))$`I` 
```

