---
title: "Use of ChatGPT"
author: "Eugene Brusilovskiy"
output: rmdformats::readthedown
---

# ChatGPT in R
In this Markdown, We will examine how to integrate R with ChatGPT. First, we will install the required R packages and let R know our OpenAI API Key.
While there are several R packages that connect to chatGPT (e.g., `chatgpt`, `gptchatteR`, etc.), many have glitches that yield error messages when the prompt is relatively long. Specifically, errors typically relate to exceeding the permitted number of tokens (currently just over 4000), which is related to the number of characters in the prompt. However, when we put in the same prompt into the chatGPT website, such an error isn't given. So the best approach to work around the issues in these apps is to submit the prompt through the website using the `httr` package, using the approach recommended here: https://rpubs.com/nirmal/setting_chat_gpt_R. 

# Getting API Key
First, we need to get an API Key for ChatGPT. According to ChatGPT itself, these are the steps to get one:

1. Sign Up or Log In to OpenAI: If you haven't already, create an account with OpenAI or log in if you already have one. You can do this by visiting the OpenAI website.
2. Navigate to the API Section: Once logged in, go to the API section of your account dashboard. This area is dedicated to API access and management.
3. Select ChatGPT API: Look for the option related to ChatGPT. OpenAI may offer various APIs, so ensure you select the one specifically for ChatGPT.
4. Generate an API Key: In the ChatGPT API section, there should be an option to generate a new API key. This process may involve agreeing to terms of service, choosing your plan (which could be free or paid, depending on OpenAI's current offerings), and possibly entering payment details.
5. Copy and Secure Your API Key: After your API key is generated, copy it and store it in a secure location. It's important to keep your API key confidential to prevent unauthorized usage.
6. Review the Documentation: Before you start using the ChatGPT API, make sure to read the provided documentation. This will help you understand the API's capabilities, limitations, usage guidelines, and how to integrate it with your applications.
7. Stay Updated: OpenAI may update their policies, pricing, or API capabilities, so it's good practice to stay informed about any changes that might affect your use of the API.


# Setting Up
```{r warning=FALSE, message=FALSE, cache=FALSE, echo=TRUE}
#knitr::opts_chunk$set(echo = TRUE)
#install.packages(c("devtools", "openai"))
#install.packages("docxtractr")
#install.packages("qdapTools")
#install.packages("chatgpt")
#library(remotes)
library(devtools)
library(tm)
library(openai)
library(docxtractr)
library(stringr)
library(qdapTools)
#library(gptchatteR)
library(data.table)
library(chatgpt)
library(httr)
library(wordcloud)
library(text)
library(tm)
library(SnowballC)
library(words)
library(NbClust)
library(stringr)
library(dplyr)
library(syuzhet)
library(textstem)
```

Here is the API Key. Get your own!
```{r warning=FALSE, message=FALSE, cache=FALSE, echo=TRUE}
my_API <- "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"
```

We proceed by setting the working directory, and importing an Excel file that contains quotes from a few books. 
```{r warning=FALSE, message=FALSE, cache=FALSE, echo=TRUE}
setwd("C:\\Users\\eugeneby\\Dropbox\\Documents\\Work and School\\Teaching\\CPLN 671 - Statistics and Data Mining\\2023_MUSA_5000_GitHub\\Data\\Lecture 23")
quotes <- read.csv(file="Book_Quotes_for_ChatGPT.csv", header=TRUE)
head(quotes)
```


# Doing Data Analysis with ChatGPT

Now, let's play around with ChatGPT! We are going to ask it to create a few variables: 1) to provide the context of each quote, 2) to let us know if it thinks the person who said the quote is named Boris, and 3) to let us know what year the quote was made.

```{r warning=FALSE, message=FALSE, cache=FALSE, echo=TRUE}
#Loading Required Libraries
library(stringr)
library(httr)
# Asking Questions to ChatGPT, Saving and Cleaning Answer
hey_chatGPT <- function(answer_my_question) {
  chat_GPT_answer <- POST(
    url = "https://api.openai.com/v1/chat/completions",
    add_headers(Authorization = paste("Bearer", my_API)),
    content_type_json(),
    encode = "json",
    body = list(
      model = "gpt-3.5-turbo-0301",
      messages = list(
        list(
          role = "user",
          content = answer_my_question
        )
      )
    )
  )
  paste(str_trim(httr::content(chat_GPT_answer)$choices[[1]]$message$content), "TOKENS USED: ", httr::content(chat_GPT_answer)$usage$total_tokens)
}


quotes$context <- NULL
for (x in 1:nrow(quotes)) {
  quotes$context[[x]] <-paste(hey_chatGPT(paste("Look at the following quote. Can you write a paragraph about the context of this quote?", quotes$Book.Quotes[[x]])))
}

quotes$boris <- NULL
for (x in 1:nrow(quotes)) {
  quotes$boris[[x]] <-paste(hey_chatGPT(paste("Look at the following quote. Do you think the person saying it is named Boris?", quotes$Book.Quotes[[x]])))
}

quotes$year <- NULL
for (x in 1:nrow(quotes)) {
  quotes$year[[x]] <-paste(hey_chatGPT(paste("Look at the following quote. In what year was this quote made, to the best of your knowledge? In your response, type the 4 digit year. If you don't know the year, leave the response blank. Here's the quote:", quotes$Book.Quotes[[x]])))
}

#Let's look at the first few rows of the data
head(quotes)
```

