---
title: "Predicting Median House Value in Philadelphia"
author: "Akira Di Sandro, Sofia Fasullo, Amy Solano"
date: "2024-10-14"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, eval = T, warning = F, message = F)

# get rid of scientific notation
options(scipen = 999)

# load appropriate packages
library(tidyverse)
library(sf)
library(FNN)
library(grid)
library(gridExtra)
library(kableExtra)
library(ggcorrplot)
library(rmarkdown)
library(pander)
library(Hmisc)
library(car)
library(DAAG)
library(MASS)
```

# Introduction

**intro paragraphs**

```{r load_data}
regdata <- read.csv("RegressionData.csv")
regshp <- st_read("Lecture 1 - RegressionData.shp/RegressionData.shp")

```

# Methods

## Data Cleaning


## Exploratory Data Analysis

[describe variables]

[formula for sample correlation coefficient]

## Multiple Regression Analysis

[specific regression equation for this problem]

[state hypothesis being tested]

## Additional Analyses

[stepwise regression]

[k-fold cross-validation]

## Software


# Results

## Exploratory Results

[describe summary stats]

```{r sum_stats}
sumstats <- regdata %>% 
  summarise(
    mean_MEDHVAL = mean(MEDHVAL),
    sd_MEDHVAL = sd(MEDHVAL),
    mean_PCTBACHMOR = mean(PCTBACHMOR),
    sd_PCTBACHMOR = sd(PCTBACHMOR),
    mean_MEDHHINC = mean(MEDHHINC),
    sd_MEDHHINC = sd(MEDHHINC),
    mean_PCTVACANT = mean(PCTVACANT),
    sd_PCTVACANT = sd(PCTVACANT),
    mean_PCTSINGLES = mean(PCTSINGLES),
    sd_PCTSINGLES = sd(PCTSINGLES),
    mean_NBELPOV100 = mean(NBELPOV100),
    sd_NBELPOV100 = sd(NBELPOV100),
    )
    
means <- sumstats %>% 
  pivot_longer(cols = all_of(starts_with("mean_")),
               names_to = "variable",
               values_to = "Mean",
               names_prefix = "mean_") %>% 
  dplyr::select(variable:Mean)


sds <- sumstats %>% 
  pivot_longer(cols = all_of(starts_with("sd_")),
               names_to = "variable",
               values_to = "SD",
               names_prefix = "sd_") %>% 
  dplyr::select(variable:SD)

# create a clean dataframe for table
sumstats_tab <- left_join(means, sds, by = "variable") %>% 
  rename(rowname = 1) %>% 
  column_to_rownames() %>% 
  t() %>% 
  data.frame() %>% 
  mutate(`Median House Value (MEDHVAL)` = paste0("$",format(round(MEDHVAL, 2), big.mark = ",", nsmall = 2)),
         `Pct Bachelor or more (PCTBACHMOR)` = paste0(format(round(PCTBACHMOR, 1), nsmall = 1), "%"),
         `Median Household Income (MEDHHINC)` = paste0("$",format(round(MEDHHINC, 2), big.mark = ",", nsmall = 2)),
         `Pct Vacant (PCTVACANT)` = paste0(format(round(PCTVACANT, 1), nsmall = 1), "%"),
         `Pct Single Unit (PCTSINGLES)` = paste0(format(round(PCTSINGLES, 1), nsmall = 1), "%"),
         `Units Below Poverty level (NBELPOV100)` = format(round(NBELPOV100), nsmall = 0),
         .keep = "unused") %>% 
  t() %>% 
  data.frame() %>% 
  rownames_to_column(var = "Variable")
 
sumstats_tab %>% 
  kbl(caption = "Table 1. Summary Statistics of Dependent and Independent Variables",
      align = "lcc") %>% 
  kable_styling() %>%
  kable_classic(full_width = F, html_font = "Cambria") 

```

[normality of variables]

```{r histograms}
# MEDHVAL
hist_medhval <- ggplot(regdata) +
  geom_histogram(aes(x = MEDHVAL), fill = "#451077FF") +
  theme_minimal() +
  labs(title = "Distribution of Median House Value",
       subtitle = "Philadelphia Block Groups",
       x = "Median House Value",
       y = "Count") +
  scale_x_continuous(labels = scales::dollar)

# PCTBACHMOR
hist_pctbach <- regdata %>% 
  mutate(PCTBACHMOR = PCTBACHMOR / 100) %>% 
  ggplot() +
  geom_histogram(aes(x = PCTBACHMOR), fill = "#451077FF") +
  theme_minimal() +
  labs(title = "Distribution of Proportion of residents with at least a Bachelor's Degree",
       subtitle = "Philadelphia Block Groups",
       x = "Proportion of residents with at least a Bachelor's Degree",
       y = "Count") +
  scale_x_continuous(labels = scales::percent)

# MEDHHINC
hist_medhhinc <- ggplot(regdata) +
  geom_histogram(aes(x = MEDHHINC), fill = "#451077FF") +
  theme_minimal() +
  labs(title = "Distribution of Median Household Income",
       subtitle = "Philadelphia Block Groups",
       x = "Median Household Income",
       y = "Count") +
  scale_x_continuous(labels = scales::dollar)

# PCTVACANT
hist_pctvac <- regdata %>% 
  mutate(PCTVACANT = PCTVACANT / 100) %>% 
  ggplot() +
  geom_histogram(aes(x = PCTVACANT), fill = "#451077FF") +
  theme_minimal() +
  labs(title = "Distribution of Proportion of housing units that are vacant",
       subtitle = "Philadelphia Block Groups",
       x = "Proportion of housing units that are vacant",
       y = "Count") +
  scale_x_continuous(labels = scales::percent)

# PCTSINGLES
hist_pctsing <- regdata %>% 
  mutate(PCTSINGLES = PCTSINGLES / 100) %>% 
  ggplot() +
  geom_histogram(aes(x = PCTSINGLES), fill = "#451077FF") +
  theme_minimal() +
  labs(title = "Distribution of Percent of housing units that are detached single family houses",
       subtitle = "Philadelphia Block Groups",
       x = "Percent of housing units that are detached single family houses",
       y = "Count") +
  scale_x_continuous(labels = scales::percent)

# NBELPOV100
hist_nbelpov <- ggplot(regdata) +
  geom_histogram(aes(x = NBELPOV100), fill = "#451077FF") +
  theme_minimal() +
  labs(title = "Distribution of Number of households living in poverty",
       subtitle = "Philadelphia Block Groups",
       caption = "Figure XX.",
       x = "Number of households living in poverty",
       y = "Count")

# log-transformed variables
regdata_log <- regdata %>% 
  mutate(LNMEDHVAL = log(MEDHVAL),
         LNPCBACHMORE = log(1 + PCTBACHMOR),
         LNNBELPOV100 = log(1 + NBELPOV100),
         LNPCTVACANT = log(1 + PCTVACANT),
         LNPCTSINGLES = log(1 + PCTSINGLES),
         LNMEDHHINC = log(MEDHHINC))

      # LNMEDHVAL
      hist_lnmedhval <- ggplot(regdata_log) +
        geom_histogram(aes(x = LNMEDHVAL), fill = "#451077FF") +
        theme_minimal() +
        labs(title = "Distribution of Log-transformed Median House Value",
             subtitle = "Philadelphia Block Groups",
             caption = "Figure XX.",
             x = "Log-transformed Median House Value",
             y = "Count")
      
      # log PCTBACHMOR
      hist_lnpctbach <- regdata_log %>% 
        ggplot() +
        geom_histogram(aes(x = LNPCBACHMORE), fill = "#451077FF") +
        theme_minimal() +
        labs(title = "Distribution of Log-transformed Proportion of residents with at least a Bachelor's Degree",
             subtitle = "Philadelphia Block Groups",
             caption = "Figure XX.",
             x = "Log-transformed Proportion of residents with at least a Bachelor's Degree",
             y = "Count")
      
      # log MEDHHINC
      hist_lnmedhhinc <- ggplot(regdata_log) +
        geom_histogram(aes(x = LNMEDHHINC), fill = "#451077FF") +
        theme_minimal() +
        labs(title = "Distribution of Log-transformed Median Household Income",
             subtitle = "Philadelphia Block Groups",
             caption = "Figure XX.",
             x = "Log-transformed Median Household Income",
             y = "Count")
      
      # log PCTVACANT
      hist_lnpctvac <- regdata_log %>% 
        ggplot() +
        geom_histogram(aes(x = LNPCTVACANT), fill = "#451077FF") +
        theme_minimal() +
        labs(title = "Distribution of Log-transformed Proportion of housing units that are vacant",
             subtitle = "Philadelphia Block Groups",
             caption = "Figure XX.",
             x = "Log-transformed Proportion of housing units that are vacant",
             y = "Count")
      
      # log PCTSINGLES
      hist_lnpctsing <- regdata_log %>% 
        ggplot() +
        geom_histogram(aes(x = LNPCTSINGLES), fill = "#451077FF") +
        theme_minimal() +
        labs(title = "Distribution of Log-transformed Percent of housing units that are detached single family houses",
             subtitle = "Philadelphia Block Groups",
             caption = "Figure XX.",
             x = "Log-transformed Percent of housing units that are detached single family houses",
             y = "Count")
      
      # log NBELPOV100
      hist_lnnbelpov <- ggplot(regdata_log) +
        geom_histogram(aes(x = LNNBELPOV100), fill = "#451077FF") +
        theme_minimal() +
        labs(title = "Distribution of Log-transformed Number of households living in poverty",
       subtitle = "Philadelphia Block Groups",
       caption = "Figure XX.",
       x = "Log-transformed Number of households living in poverty",
       y = "Count")


```



Note that the dependent variable does look more or less normal after the transformation – hence, LNMEDHVAL will be used as the dependent variable in the regression analysis. You will also see that for the predictors, the logarithmic transformation only helps normalize the NBELPOV100 variable (so we will use LNNBELPOV100 in the subsequent analyses). The other variables have a large spike at zero (i.e., zero-inflated distributions) after the transformations, so we will use the original, untransformed PCBACHMORE, PCTVACANT, and PCTSINGLES variables in the regression.

```{r}
ggplot(regdata) + 
  geom_point(aes(x=LNMEDHVAL, y=LNNBELPOV100))

ggplot(regdata) + 
  geom_point(aes(x=LNMEDHVAL, y=PCTBACHMOR))

ggplot(regdata) + 
  geom_point(aes(x=LNMEDHVAL, y=PCTVACANT))

ggplot(regdata) + 
  geom_point(aes(x=LNMEDHVAL, y=PCTSINGLES))
```

```{r}
predictors <- regdata %>% select(LNNBELPOV100,PCTVACANT,PCTSINGLES,PCTBACHMOR)

rcorr(as.matrix(predictors), type="pearson") 
```


```{r}
ggcorrplot(cor(predictors), lab=TRUE)
```


```{r}
fit_data <- lm(LNMEDHVAL ~ LNNBELPOV100 + PCTVACANT + PCTSINGLES + PCTBACHMOR, data=regdata)
summary(fit_data)
```
```{r}
anova(fit_data)
```

```{r}
vif(fit_data)
```

```{r}
predicted <- fitted(fit_data)
resid <- residuals(fit_data)
resid_std <- rstandard(fit_data)
```


```{r}
ggplot(regshp) +
      geom_sf(aes(fill = LNMEDHVAL))+
      scale_fill_distiller(palette="YlGnBu")+
      labs(title = "Median House Values, log transformed")+
      theme_void()
```


```{r}
MapVacant <- ggplot(regshp) +
      geom_sf(aes(fill = PCTVACANT))+
      scale_fill_distiller(palette="YlGnBu")+
      labs(title = "Percent of housing units that are vacant")+
      theme_void()

MapSingles <- ggplot(regshp) +
      geom_sf(aes(fill = PCTSINGLES))+
      scale_fill_distiller(palette="YlGnBu")+
      labs(title = "Percent single homes")+
      theme_void()

BachMap <- ggplot(regshp) +
      geom_sf(aes(fill = PCTBACHMOR))+
      scale_fill_distiller(palette="YlGnBu")+
      labs(title = "Percent with bachelor's degree or higher")+
      theme_void()

PovMap <- ggplot(regshp) +
      geom_sf(aes(fill = LNNBELPOV))+
      scale_fill_distiller(palette="YlGnBu")+
      labs(title = "Households in poverty, log transformed")+
      theme_void()

grid.arrange(MapVacant, MapSingles, BachMap, PovMap, nrow=2)
```

```{r}
plot(x=predicted, y=resid_std)
```

```{r}
hist(resid)
```


```{r}
stepm <- stepAIC(fit_data, direction="both")
stepm$anova
```


```{r}
cv <- CVlm(data=regdata, form.lm=fit_data, m=5)				        #m=5 sets it to 5 folds
#Extracting MSEs
mse <- attr(cv, "ms")
rmse <- sqrt(mse)					                      #Obtaining RMSE for model 1
```






```{r}
rmse <- sqrt(mse)						  
rmse
```



```{r}
fit2 <- lm(LNMEDHVAL ~ PCTVACANT + MEDHHINC, data=regdata)

cv2 <- CVlm(data=regdata, fit2, m=5)				        #m=5 sets it to 5 folds
#Extracting MSEs
mse2 <- attr(cv2, "ms")
rmse2 <- sqrt(mse2)					                      #Obtaining RMSE for model 2
rmse2
```

```{r}
rmse2
```



```{r}
anova(fit_data, fit2)
```


```{r}
step2 <- stepAIC(fit2, direction="both")
step2$anova
```

```{r}
anova(stepm, step2)
```

